---
title: "MUSIC analysis on CD8+ T Cell (stimulated and unstimulated) CROP-seq data"
author: "Kaixuan Luo"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(results = "hold", fig.align = "center", message=FALSE, warning=FALSE)
```

MUSIC website: https://github.com/bm2-lab/MUSIC

## About the data sets

CROP-seq datasets: `/project2/xinhe/yifan/Factor_analysis/shared_data/`.
The data are Seurat objects, with raw gene counts stored in `obj@assays$RNA@counts`, 
and cell meta data stored in `obj@meta.data`.
Normalized and scaled data used for GSFA are stored in `obj@assays$RNA@scale.data`, 
the rownames of which are the 6k genes used for GSFA.

Load packages
```{r load-packages, message=FALSE, warning=FALSE}
dyn.load('/software/geos-3.7.0-el7-x86_64/lib64/libgeos_c.so') # attach the geos lib for Seurat
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(MUSIC))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(ggplot2))
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 13),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)

source("R/plotting_functions.R")
```

Functions
```{r functions}
## Adapted over MUSIC's Diff_topic_distri() function
Empirical_topic_prob_diff <- function(model, perturb_information,
                                      permNum = 10^4, seed = 1000){
  require(reshape2)
  require(dplyr)
  require(ComplexHeatmap)
  options(warn = -1)
  prob_mat <- model@gamma
  row.names(prob_mat) <- model@documents
  topicNum <- ncol(prob_mat)
  topicName <- paste0('Topic_', 1:topicNum)
  colnames(prob_mat) <- topicName
  ko_name <- unique(perturb_information)
  prob_df <- data.frame(prob_mat, 
                        samples = rownames(prob_mat),
                        knockout = perturb_information)
  
  prob_df <- melt(prob_df, id = c('samples', 'knockout'), variable.name = "topic")
  
  summary_df <- prob_df %>%
    group_by(knockout, topic) %>%
    summarise(number = sum(value)) %>%
    ungroup() %>%
    group_by(knockout) %>%
    mutate(cellNum = sum(number)) %>%
    ungroup() %>%
    mutate(ratio = number/cellNum)
  
  summary_df$ctrlNum <- rep(summary_df$cellNum[summary_df$knockout == "CTRL"],
                            length(ko_name))
  summary_df$ctrl_ratio <- rep(summary_df$ratio[summary_df$knockout == "CTRL"],
                               length(ko_name))
  summary_df <- summary_df %>% mutate(diff_index = ratio - ctrl_ratio)
  
  test_df <- data.frame(matrix(nrow = length(ko_name) * topicNum, ncol = 5))
  colnames(test_df) <- c("knockout", "topic", "obs_t_stats", "obs_pval", "empirical_pval")
  k <- 1
  for(i in topicName){
    prob_df.topic <- prob_df[prob_df$topic == i, ]
    ctrl_topic <- prob_df.topic$value[prob_df.topic$knockout == "CTRL"]
    ctrl_topic_z <- (ctrl_topic - mean(ctrl_topic)) / sqrt(var(ctrl_topic))
    for(j in ko_name){
      ko_topic <- prob_df.topic$value[prob_df.topic$knockout == j]
      ko_topic_z <- (ko_topic - mean(ctrl_topic)) / sqrt(var(ctrl_topic))
      test_df$knockout[k] <- j
      test_df$topic[k] <- i
      test <- t.test(ko_topic_z, ctrl_topic_z)
      test_df$obs_t_stats[k] <- test$statistic
      test_df$obs_pval[k] <- test$p.value
      k <- k + 1
    }
  }
  
  ## Permutation on the perturbation conditions:
  # permNum <- 10^4
  print(paste0("Performing permutation for ", permNum, " rounds."))
  perm_t_stats <- matrix(0, nrow = nrow(test_df), ncol = permNum)
  set.seed(seed)
  for (perm in 1:permNum){
    perm_prob_df <- data.frame(prob_mat, 
                               samples = rownames(prob_mat),
                               knockout = perturb_information[sample(length(perturb_information))])
    perm_prob_df <- melt(perm_prob_df, id = c('samples', 'knockout'), variable.name = "topic")
    k <- 1
    for(i in topicName){
      perm_prob_df.topic <- perm_prob_df[perm_prob_df$topic == i, ]
      ctrl_topic <- perm_prob_df.topic$value[perm_prob_df.topic$knockout == "CTRL"]
      ctrl_topic_z <- (ctrl_topic - mean(ctrl_topic)) / sqrt(var(ctrl_topic))
      for(j in ko_name){
        ko_topic <- perm_prob_df.topic$value[perm_prob_df.topic$knockout == j]
        ko_topic_z <- (ko_topic - mean(ctrl_topic)) / sqrt(var(ctrl_topic))
        test <- t.test(ko_topic_z, ctrl_topic_z)
        perm_t_stats[k, perm] <- test$statistic
        k <- k + 1
      }
    }
    if (perm %% 1000 == 0){
      print(paste0(perm, " rounds finished."))
    }
  }
  ## Compute two-sided empirical p value:
  for (k in 1:nrow(test_df)){
    test_df$empirical_pval[k] <-
      2 * min(mean(perm_t_stats[k, ] <= test_df$obs_t_stats[k]),
              mean(perm_t_stats[k, ] >= test_df$obs_t_stats[k]))
  }
  test_df <- test_df %>%
    mutate(empirical_pval = ifelse(empirical_pval == 0, 1/permNum, empirical_pval)) %>%
    mutate(empirical_pval = ifelse(empirical_pval > 1, 1, empirical_pval))
  
  summary_df <- inner_join(summary_df, test_df, by = c("knockout", "topic"))
  summary_df <- summary_df %>%
    mutate(polar_log10_pval = ifelse(obs_t_stats > 0, -log10(empirical_pval), log10(empirical_pval)))
  return(summary_df)
}

```

Set directories
```{r set-dir}
data_dir <- "/project2/xinhe/yifan/Factor_analysis/Stimulated_T_Cells/"
dir.create("/project2/xinhe/kevinluo/GSFA/music_analysis/Stimulated_T_Cells", recursive = TRUE, showWarnings = FALSE)
```

Load the T Cells CROP-seq data
```{r load-data}
combined_obj <- readRDS('/project2/xinhe/yifan/Factor_analysis/shared_data/TCells_cropseq_data_seurat.rds')
```

Separate stimulated and unstimulated cells into two data sets, and run those separately.
```{r cell-covariate-matrix}
metadata <- combined_obj@meta.data
metadata[1:5, ]

table(metadata$orig.ident)

stimulated_cells <- rownames(metadata)[which(endsWith(metadata$orig.ident, "S"))]
cat(length(stimulated_cells), "stimulated cells. \n")
unstimulated_cells <- rownames(metadata)[which(endsWith(metadata$orig.ident, "N"))]
cat(length(unstimulated_cells), "unstimulated cells. \n")
```

## Run simulated data
```{r}
dir.create("/project2/xinhe/kevinluo/GSFA/music_analysis/Stimulated_T_Cells/stimulated", recursive = TRUE, showWarnings = FALSE)
res_dir <- "/project2/xinhe/kevinluo/GSFA/music_analysis/Stimulated_T_Cells/stimulated"
dir.create(file.path(res_dir,"/music_output"), recursive = TRUE, showWarnings = FALSE)
```

```{r, eval=FALSE}
setwd(res_dir)
```

### 0. Load input data

```{r load_input_data}
feature.names <- data.frame(fread(paste0(data_dir, "GSE119450_RAW/D1N/genes.tsv"),
                                  header = FALSE), stringsAsFactors = FALSE)

expression_profile <- combined_obj@assays$RNA@counts[, stimulated_cells]
rownames(expression_profile) <- feature.names$V2[match(rownames(expression_profile),
                                                       feature.names$V1)]
cat("Dimension of expression profile matrix: \n")
dim(expression_profile)

targets <- names(combined_obj@meta.data)[4:24]
targets[targets == "NonTarget"] <- "CTRL"
cat("Targets: \n")
print(targets)

perturb_information <- apply(combined_obj@meta.data[stimulated_cells, 4:24], 1,
                             function(x){ targets[which(x > 0)] })
```

### 1. Data preprocessing

```{r preprocess, eval=FALSE}
crop_seq_list <- Input_preprocess(expression_profile, perturb_information)

crop_seq_qc <- Cell_qc(crop_seq_list$expression_profile,
                       crop_seq_list$perturb_information,
                       species = "Hs", plot = F)
saveRDS(crop_seq_qc, "music_output/music_crop_seq_qc.rds")

crop_seq_imputation <- Data_imputation(crop_seq_qc$expression_profile,
                                       crop_seq_qc$perturb_information,
                                       cpu_num = 5)
saveRDS(crop_seq_imputation, "music_output/music_imputation.merged.rds")

crop_seq_filtered <- Cell_filtering(crop_seq_imputation$expression_profile,
                                    crop_seq_imputation$perturb_information,
                                    cpu_num = 5)
saveRDS(crop_seq_filtered, "music_output/music_filtered.merged.rds")
```

### 2. Model building
```{r topics_models, eval=FALSE}
crop_seq_vargene <- Get_high_varGenes(crop_seq_filtered$expression_profile,
                                      crop_seq_filtered$perturb_information, plot = T)
saveRDS(crop_seq_vargene, "music_output/music_vargene.merged.rds")

crop_seq_vargene <- readRDS("music_output/music_vargene.merged.rds")

## Get_topics() can take up to a few hours to finish, 
## depending on the size of data
system.time(
  topic_1 <- Get_topics(crop_seq_vargene$expression_profile,
                        crop_seq_vargene$perturb_information,
                        topic_number = 5))
saveRDS(topic_1, "music_output/music_merged_5_topics.rds")

system.time(
  topic_2 <- Get_topics(crop_seq_vargene$expression_profile,
                        crop_seq_vargene$perturb_information,
                        topic_number = 10))
saveRDS(topic_2, "music_output/music_merged_10_topics.rds")

system.time(
  topic_3 <- Get_topics(crop_seq_vargene$expression_profile,
                        crop_seq_vargene$perturb_information,
                        topic_number = 15))
saveRDS(topic_3, "music_output/music_merged_15_topics.rds")

system.time(
  topic_4 <- Get_topics(crop_seq_vargene$expression_profile,
                        crop_seq_vargene$perturb_information,
                        topic_number = 20))
saveRDS(topic_4, "music_output/music_merged_20_topics.rds")

## try fewer numbers of topics
system.time(
  topic_5 <- Get_topics(crop_seq_vargene$expression_profile,
                        crop_seq_vargene$perturb_information,
                        topic_number = 4))
saveRDS(topic_5, "music_output/music_merged_4_topics.rds")

system.time(
  topic_6 <- Get_topics(crop_seq_vargene$expression_profile,
                        crop_seq_vargene$perturb_information,
                        topic_number = 6))
saveRDS(topic_6, "music_output/music_merged_6_topics.rds")
```

### 3. Pick the number of topics
```{r select_num_topics, eval=FALSE}
topic_1 <- readRDS("music_output/music_merged_4_topics.rds")
topic_2 <- readRDS("music_output/music_merged_5_topics.rds")
topic_3 <- readRDS("music_output/music_merged_6_topics.rds")
topic_4 <- readRDS("music_output/music_merged_10_topics.rds")
topic_5 <- readRDS("music_output/music_merged_15_topics.rds")
topic_6 <- readRDS("music_output/music_merged_20_topics.rds")

topic_model_list <- list()
topic_model_list$models <- list()
topic_model_list$perturb_information <- topic_1$perturb_information
topic_model_list$models[[1]] <- topic_1$models[[1]]
topic_model_list$models[[2]] <- topic_2$models[[1]]
topic_model_list$models[[3]] <- topic_3$models[[1]]
topic_model_list$models[[4]] <- topic_4$models[[1]]
topic_model_list$models[[5]] <- topic_5$models[[1]]
topic_model_list$models[[6]] <- topic_6$models[[1]]

optimalModel <- Select_topic_number(topic_model_list$models,
                                    plot = T,
                                    plot_path = "music_output/select_topic_number_4to6to20.pdf")
```

## Summarize the results

### Summarize the results under 20 topics to be comparable to GSFA

Gene ontology annotations for top topics
```{r GO_20topics, eval=FALSE}
topic_res <- readRDS("music_output/music_merged_20_topics.rds")

topic_func <- Topic_func_anno(topic_res$models[[1]], species = "Hs")
saveRDS(topic_func, "music_output/topic_func_20_topics.rds")
```

```{r GO_20topics-2, eval=FALSE, fig.weight=14, fig.height=12}
topic_func <- readRDS("music_output/topic_func_20_topics.rds")
pdf("music_output/music_merged_20_topics_GO_annotations.pdf",
    width = 14, height = 12)
ggplot(topic_func$topic_annotation_result) +
  geom_point(aes(x = Cluster, y = Description,
                 size = Count, color = -log10(qvalue))) +
  scale_color_gradientn(colors = c("blue", "red")) +
  theme_bw() +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

Perturbation effect prioritizing
```{r Diff_topic_distri_20topics, eval=FALSE}
# calculate topic distribution for each cell.
distri_diff <- Diff_topic_distri(topic_res$models[[1]],
                                 topic_res$perturb_information,
                                 plot = T)
# saveRDS(distri_diff, "music_output/distri_diff_20_topics.rds")

distri_diff <- readRDS("music_output/distri_diff_20_topics.rds")
t_D_diff_matrix <- dcast(distri_diff %>% dplyr::select(knockout, variable, t_D_diff),
                         knockout ~ variable)
rownames(t_D_diff_matrix) <- t_D_diff_matrix$knockout
t_D_diff_matrix$knockout <- NULL
```

```{r t_D_diff_matrix_20topics, eval=FALSE, fig.width=12, fig.height=8}
pdf("music_output/music_merged_20_topics_TPD_heatmap.pdf", width = 12, height = 8)
Heatmap(t_D_diff_matrix,
        name = "Topic probability difference (vs ctrl)",
        cluster_rows = T, cluster_columns = T,
        column_names_rot = 45,
        heatmap_legend_param = list(title_gp = gpar(fontsize = 12, fontface = "bold")))
dev.off()
```

Calculate the overall perturbation effect ranking list without "offTarget_Info".
```{r Rank_overall_20topics}
distri_diff <- readRDS(file.path(res_dir, "music_output/distri_diff_20_topics.rds"))

rank_overall_result <- Rank_overall(distri_diff)
print(rank_overall_result)
# saveRDS(rank_overall_result, "music_output/rank_overall_result_20_topics.rds")
```

calculate the topic-specific ranking list.
```{r Rank_specific_20topics}
rank_topic_specific_result <- Rank_specific(distri_diff)
head(rank_topic_specific_result, 10)
# saveRDS(rank_topic_specific_result, "music_output/rank_topic_specific_result_20_topics.rds")
```

calculate the perturbation correlation.
```{r Correlation_perturbation_20topics}
perturb_cor <- Correlation_perturbation(distri_diff,
                                        cutoff = 0.5, gene = "all", plot = T,
                                        plot_path = file.path(res_dir, "music_output/correlation_network_20_topics.pdf"))

head(perturb_cor, 10)
# saveRDS(perturb_cor, "music_output/perturb_cor_20_topics.rds")
```

Adaptation to the code to generate calibrated empirical TPD scores
```{r Empirical_topic_prob_diff_20topics, eval=FALSE}
summary_df <- Empirical_topic_prob_diff(topic_res$models[[1]],
                                        topic_res$perturb_information)
saveRDS(summary_df, "music_output/music_merged_20_topics_ttest_summary.rds")
```


```{r load_summary_df_20topics}
summary_df <- readRDS(file.path(res_dir, "music_output/music_merged_20_topics_ttest_summary.rds"))

summary_df$topic <- gsub("_", " ", summary_df$topic)
summary_df$topic <- factor(summary_df$topic, levels = paste("Topic", 1:length(unique(summary_df$topic))))

summary_df$fdr <- p.adjust(summary_df$empirical_pval, method = "BH")
summary_df$bonferroni_adj <- p.adjust(summary_df$empirical_pval, method = "bonferroni")

log10_pval_mat <- dcast(summary_df %>% dplyr::select(knockout, topic, polar_log10_pval),
                        knockout ~ topic)
rownames(log10_pval_mat) <- log10_pval_mat$knockout
log10_pval_mat$knockout <- NULL

effect_mat <- dcast(summary_df %>% dplyr::select(knockout, topic, obs_t_stats), knockout ~ topic)
rownames(effect_mat) <- effect_mat$knockout
effect_mat$knockout <- NULL

fdr_mat <- dcast(summary_df %>% dplyr::select(knockout, topic, fdr), knockout ~ topic)
rownames(fdr_mat) <- fdr_mat$knockout
fdr_mat$knockout <- NULL

bonferroni_mat <- dcast(summary_df %>% dplyr::select(knockout, topic, bonferroni_adj), knockout ~ topic)
rownames(bonferroni_mat) <- bonferroni_mat$knockout
bonferroni_mat$knockout <- NULL
```

```{r empirical_tstats_heatmap_20topics, fig.width=12, fig.height=8}
# pdf("music_output/music_merged_20_topics_empirical_tstats_heatmap.pdf",
#     width = 12, height = 8)
ht <- Heatmap(log10_pval_mat,
              name = "Polarized empirical t-test -log10(p-value)\n(KO vs ctrl cell topic probs)",
              col = circlize::colorRamp2(breaks = c(-4, 0, 4), colors = c("blue", "grey90", "red")),
              cluster_rows = T, cluster_columns = T,
              column_names_rot = 45,
              heatmap_legend_param = list(title_gp = gpar(fontsize = 12,
                                                          fontface = "bold")))
draw(ht)
# dev.off()
```

```{r effect_fdr_dotplot_20topics, dev=c('png', 'pdf'), fig.width=10, fig.height=6}
# pdf("music_output/music_merged_20_topics_empirical_tstats_fdr_dotplot.pdf",
#     width = 12, height = 8)
KO_names <- rownames(fdr_mat)
dotplot_effectsize(t(effect_mat), t(fdr_mat),
                   reorder_markers = c(KO_names[KO_names!="CTRL"], "CTRL"),
                   color_lgd_title = "MUSIC T statistics",
                   size_lgd_title = "FDR",
                   max_score = 20,
                   min_score = -20,
                   by_score = 10) + 
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1))
# dev.off()
```

```{r effect_bonferroni_dotplot_20topics, dev=c('png', 'pdf'), fig.width=10, fig.height=6}
KO_names <- rownames(bonferroni_mat)
dotplot_effectsize(t(effect_mat), t(bonferroni_mat),
                   reorder_markers = c(KO_names[KO_names!="CTRL"], "CTRL"),
                   color_lgd_title = "MUSIC T statistics",
                   size_lgd_title = "Bonferroni\nadjusted p-value",
                   max_score = 20,
                   min_score = -20,
                   by_score = 10) + 
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

### Summarize the results using the optimal number of topics selected by the score
```{r choose_optimal_num_topics, eval=FALSE}
topic_1 <- readRDS("music_output/music_merged_4_topics.rds")
topic_2 <- readRDS("music_output/music_merged_5_topics.rds")
topic_3 <- readRDS("music_output/music_merged_6_topics.rds")

topic_model_list <- list()
topic_model_list$models <- list()
topic_model_list$perturb_information <- topic_1$perturb_information
topic_model_list$models[[1]] <- topic_1$models[[1]]
topic_model_list$models[[2]] <- topic_2$models[[1]]
topic_model_list$models[[3]] <- topic_3$models[[1]]

optimalModel <- Select_topic_number(topic_model_list$models,
                                    plot = T,
                                    plot_path = "music_output/select_topic_number_4to6.pdf")
optimalModel
saveRDS(optimalModel, "music_output/optimalModel_4_topics.rds")
```

Gene ontology annotations for top topics
```{r GO_4topics, eval=FALSE}
topic_func <- Topic_func_anno(optimalModel, species = "Hs", plot_path = "music_output/topic_annotation_GO_4_topics.pdf")
saveRDS(topic_func, "music_output/topic_func_4_topics.rds")
```

```{r GO_4topics-2, eval=FALSE, fig.weight=14, fig.height=12}
topic_func <- readRDS("music_output/topic_func_4_topics.rds")
pdf("music_output/music_merged_4_topics_GO_annotations.pdf",
    width = 14, height = 12)
ggplot(topic_func$topic_annotation_result) +
  geom_point(aes(x = Cluster, y = Description,
                 size = Count, color = -log10(qvalue))) +
  scale_color_gradientn(colors = c("blue", "red")) +
  theme_bw() +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

Perturbation effect prioritizing
```{r Diff_topic_distri_4topics, eval=FALSE}
# calculate topic distribution for each cell.
distri_diff <- Diff_topic_distri(optimalModel,
                                 topic_model_list$perturb_information,
                                 plot = T,
                                 plot_path = "music_output/distribution_of_topic_4_topics.pdf")
saveRDS(distri_diff, "music_output/distri_diff_4_topics.rds")

t_D_diff_matrix <- dcast(distri_diff %>% dplyr::select(knockout, variable, t_D_diff),
                         knockout ~ variable)
rownames(t_D_diff_matrix) <- t_D_diff_matrix$knockout
t_D_diff_matrix$knockout <- NULL

```

```{r t_D_diff_matrix_4topics, eval=FALSE, fig.width=12, fig.height=8}
pdf("music_output/music_merged_4_topics_TPD_heatmap.pdf", width = 12, height = 8)
Heatmap(t_D_diff_matrix,
        name = "Topic probability difference (vs ctrl)",
        cluster_rows = T, cluster_columns = T,
        column_names_rot = 45,
        heatmap_legend_param = list(title_gp = gpar(fontsize = 12, fontface = "bold")))
dev.off()
```

The overall perturbation effect ranking list.
```{r Rank_overall_4topics}
distri_diff <- readRDS(file.path(res_dir, "music_output/distri_diff_4_topics.rds"))

rank_overall_result <- Rank_overall(distri_diff)
# print(rank_overall_result)
# saveRDS(rank_overall_result, "music_output/rank_overall_4_topics_result.rds")
```

Topic-specific ranking list.
```{r Rank_specific_4topics}
rank_topic_specific_result <- Rank_specific(distri_diff)
# print(rank_topic_specific_result)
# saveRDS(rank_topic_specific_result, "music_output/rank_topic_specific_4_topics_result.rds")
```

Perturbation correlation.
```{r Correlation_perturbation_4topics}
perturb_cor <- Correlation_perturbation(distri_diff,
                                        cutoff = 0.5, gene = "all", plot = T,
                                        plot_path = file.path(res_dir, "music_output/correlation_network_4_topics.pdf"))

# head(perturb_cor, 10)
# saveRDS(perturb_cor, "music_output/perturb_cor_4_topics.rds")
```

Adaptation to the code to generate calibrated empirical TPD scores

```{r Empirical_topic_prob_diff_4topics, eval=FALSE}
summary_df <- Empirical_topic_prob_diff(optimalModel,
                                        topic_model_list$perturb_information)
saveRDS(summary_df, "music_output/music_merged_4_topics_ttest_summary.rds")
```

```{r load_summary_df_4topics}
summary_df <- readRDS(file.path(res_dir, "music_output/music_merged_4_topics_ttest_summary.rds"))

summary_df$topic <- gsub("_", " ", summary_df$topic)
summary_df$topic <- factor(summary_df$topic, levels = paste("Topic", 1:length(unique(summary_df$topic))))

summary_df$fdr <- p.adjust(summary_df$empirical_pval, method = "BH")
summary_df$bonferroni_adj <- p.adjust(summary_df$empirical_pval, method = "bonferroni")

log10_pval_mat <- dcast(summary_df %>% dplyr::select(knockout, topic, polar_log10_pval),
                        knockout ~ topic)
rownames(log10_pval_mat) <- log10_pval_mat$knockout
log10_pval_mat$knockout <- NULL

effect_mat <- dcast(summary_df %>% dplyr::select(knockout, topic, obs_t_stats), knockout ~ topic)
rownames(effect_mat) <- effect_mat$knockout
effect_mat$knockout <- NULL

fdr_mat <- dcast(summary_df %>% dplyr::select(knockout, topic, fdr), knockout ~ topic)
rownames(fdr_mat) <- fdr_mat$knockout
fdr_mat$knockout <- NULL

bonferroni_mat <- dcast(summary_df %>% dplyr::select(knockout, topic, bonferroni_adj), knockout ~ topic)
rownames(bonferroni_mat) <- bonferroni_mat$knockout
bonferroni_mat$knockout <- NULL
```

```{r empirical_tstats_heatmap_4topics, eval=TRUE, fig.width=12, fig.height=8}
# pdf("music_output/music_merged_4_topics_empirical_tstats_heatmap.pdf",
#     width = 12, height = 8)
ht <- Heatmap(log10_pval_mat,
              name = "Polarized empirical t-test -log10(p-value)\n(KO vs ctrl cell topic probs)",
              col = circlize::colorRamp2(breaks = c(-4, 0, 4), colors = c("blue", "grey90", "red")),
              cluster_rows = T, cluster_columns = T,
              column_names_rot = 45,
              heatmap_legend_param = list(title_gp = gpar(fontsize = 12,
                                                          fontface = "bold")))
draw(ht)
# dev.off()
```

```{r effect_fdr_dotplot_4topics, dev=c('png', 'pdf'),  fig.width=4, fig.height=5.5}
# pdf("music_output/music_merged_4_topics_empirical_tstats_fdr_dotplot.pdf",
#     width = 12, height = 8)
KO_names <- rownames(fdr_mat)
dotplot_effectsize(t(effect_mat), t(fdr_mat),
                   reorder_markers = c(KO_names[KO_names!="CTRL"], "CTRL"),
                   color_lgd_title = "MUSIC T statistics",
                   size_lgd_title = "FDR",
                   max_score = 20,
                   min_score = -20,
                   by_score = 10) + 
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1))
# dev.off()
```


```{r effect_bonferroni_dotplot_4topics, dev=c('png', 'pdf'),  fig.width=4, fig.height=5.5}
KO_names <- rownames(bonferroni_mat)
dotplot_effectsize(t(effect_mat), t(bonferroni_mat),
                   reorder_markers = c(KO_names[KO_names!="CTRL"], "CTRL"),
                   color_lgd_title = "MUSIC T statistics",
                   size_lgd_title = "Bonferroni\nadjusted p-value",
                   max_score = 20,
                   min_score = -20,
                   by_score = 10) + 
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1))
```
