---
title: "Run SCEPTRE on CD8+ T Cell (stimulated and unstimulated) CROP-seq data"
author: "Kaixuan Luo"
date: "2022-05-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

## About the data sets

CROP-seq datasets: `/project2/xinhe/yifan/Factor_analysis/shared_data/`

The data are Seurat objects, with raw gene counts stored in `obj@assays$RNA@counts`, 
and cell meta data stored in `obj@meta.data`.
Normalized and scaled data used for GSFA are stored in `obj@assays$RNA@scale.data`, 
the rownames of which are the 6k genes used for GSFA.

## Analysis

Scripts for running the analysis: `code/sceptre_analysis.R`, `code/sceptre_permute_analysis.R`
```{bash, include=FALSE, eval=FALSE}
# T cells stimulated data
sbatch --mem=50G --cpus-per-task=10 ~/projects/GSFA_analysis/code/run_sceptre_Tcells_stimulated_data.sbatch

# T cells unstimulated data
sbatch --mem=50G --cpus-per-task=10 ~/projects/GSFA_analysis/code/run_sceptre_Tcells_unstimulated_data.sbatch
```

```{bash, include=FALSE, eval=FALSE}
# Permutation analysis of T cells stimulated data
for i in {1..10}
do
  sbatch --mem=50G --cpus-per-task=10 ~/projects/GSFA_analysis/code/run_sceptre_permuted_data.sbatch \
  /project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data_updated_covariates/simulated_data  i
done
```

Load packages
```{r, message=FALSE, warning=FALSE}
dyn.load('/software/geos-3.7.0-el7-x86_64/lib64/libgeos_c.so') # attach the geos lib for Seurat
suppressPackageStartupMessages(library(tidyverse))
library(cowplot)
library(Matrix)
library(sceptre)
library(Seurat)
source("code/plotting_functions.R")
```

### Prepare input data for SCEPTRE

Load the T Cells CROP-seq data
```{r load-data}
TCells_data <- readRDS('/project2/xinhe/yifan/Factor_analysis/shared_data/TCells_cropseq_data_seurat.rds')
```

We first prepare three separate data objects required to run SCEPTRE: 
the gene expression matrix, the perturbation matrix, and the covariate matrix.

Separate stimulated and unstimulated cells into two data sets, and run those separately.
```{r cell-covariate-matrix}
metadata <- TCells_data@meta.data
metadata[1:5, ]

table(metadata$orig.ident)

stimulated_cells <- rownames(metadata)[which(endsWith(metadata$orig.ident, "S"))]
cat(length(stimulated_cells), "stimulated cells. \n")
unstimulated_cells <- rownames(metadata)[which(endsWith(metadata$orig.ident, "N"))]
cat(length(unstimulated_cells), "unstimulated cells. \n")
```

#### Stimulated T cells

We first prepare three separate data objects required to run SCEPTRE: 
the gene expression matrix, the perturbation matrix, and the covariate matrix.

1. Gene expression matrices
Gene expression (gene-by-cell) raw count matrix
```{r stimulated-exp-matrix}
gene_matrix <- TCells_data@assays$RNA@counts[, stimulated_cells]
# gene-by-cell expression matrix
gene_matrix[1:5, 1:3]
dim(gene_matrix)
```

2. Covariate matrix
```{r stimulated-covariate-matrix}
# covariate_matrix <- metadata[stimulated_cells, c('nCount_RNA', 'nFeature_RNA', 'percent_mt', 'gRNA_umi_count')]
covariate_matrix <- metadata[stimulated_cells, c('nCount_RNA', 'nFeature_RNA', 'percent_mt')]
covariate_matrix[1:5,]
dim(covariate_matrix)
```

3. Perturbation matrix (a binary matrix of perturbations,  rows are gRNA groups and columns are cell barcodes)
```{r stimulated-perturbation-matrix}
combined_perturbation_matrix <- t(metadata[stimulated_cells, 4:24])
combined_perturbation_matrix[1:5,1:3]
dim(combined_perturbation_matrix)
range(combined_perturbation_matrix)
```

Specify the gene-gRNA group pairs to test for association

We include the genes used for GSFA in this analysis
```{r}
# Normalized and scaled data used for GSFA, the rownames of which are the 6k genes used for GSFA
scaled_gene_matrix <- TCells_data@assays$RNA@scale.data
dim(scaled_gene_matrix)
selected_gene_ids <- rownames(scaled_gene_matrix)

gRNA_groups <- rownames(combined_perturbation_matrix)
pairs <- expand.grid(selected_gene_ids, gRNA_groups)
gene_gRNA_group_pairs <- data.frame(gene_id = pairs$Var1, gRNA_group = pairs$Var2, pair_type = "candidate")
gene_gRNA_group_pairs[gene_gRNA_group_pairs$gRNA_group == "NonTarget", "pair_type"] <- "negative_control"
table(gene_gRNA_group_pairs$pair_type)
table(gene_gRNA_group_pairs$gRNA_group)
dim(gene_gRNA_group_pairs)
```

```{r, eval=FALSE}
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data_updated_covariates/simulated_data'
if(!dir.exists(datadir)) dir.create(datadir, recursive = TRUE)
save(list = c("gene_matrix", "combined_perturbation_matrix", "covariate_matrix"),
     file = file.path(datadir, 'data.matrices.RData'))
saveRDS(gene_gRNA_group_pairs, file.path(datadir, "gene.gRNA.group.pairs.rds"))
```


#### Unstimulated T cells

1. Gene expression matrices
Gene expression (gene-by-cell) raw count matrix
```{r unstimulated-exp-matrix}
gene_matrix <- TCells_data@assays$RNA@counts[, unstimulated_cells]
# gene-by-cell expression matrix
gene_matrix[1:5, 1:3]
dim(gene_matrix)
```

2. Covariate matrix
```{r unstimulated-covariate-matrix}
# covariate_matrix <- metadata[unstimulated_cells, c('nCount_RNA', 'nFeature_RNA', 'percent_mt', 'gRNA_umi_count')]
covariate_matrix <- metadata[unstimulated_cells, c('nCount_RNA', 'nFeature_RNA', 'percent_mt')]
covariate_matrix[1:5,]
dim(covariate_matrix)
```

3. Perturbation matrix (a binary matrix of perturbations,  rows are gRNA groups and columns are cell barcodes)
```{r unstimulated-perturbation-matrix}
combined_perturbation_matrix <- t(metadata[unstimulated_cells, 4:24])
combined_perturbation_matrix[1:5,1:3]
dim(combined_perturbation_matrix)
range(combined_perturbation_matrix)
```


```{r, eval=FALSE}
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data_updated_covariates/unsimulated_data'
if(!dir.exists(datadir)) dir.create(datadir, recursive = TRUE)
save(list = c("gene_matrix", "combined_perturbation_matrix", "covariate_matrix"),
     file = file.path(datadir, 'data.matrices.RData'))
saveRDS(gene_gRNA_group_pairs, file.path(datadir, "gene.gRNA.group.pairs.rds"))
```


### Run SCEPTRE

#### Stimulated T cells
```{r load-stimulated-input-data}
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/simulated_data'
load(file.path(datadir, "data.matrices.RData"))
gene_gRNA_group_pairs <- readRDS(file.path(datadir, "gene.gRNA.group.pairs.rds"))

cat(sprintf('Dimenstion of gene expression matrix: %d rows %d columns.\n', nrow(gene_matrix), ncol(gene_matrix)))
cat(sprintf('Dimenstion of combined perturbation matrix: %d rows %d columns.\n', nrow(combined_perturbation_matrix), ncol(combined_perturbation_matrix)))
cat(sprintf('Dimenstion of covariate matrix: %d rows %d columns.\n', nrow(covariate_matrix), ncol(covariate_matrix)))
cat(sprintf('Dimenstion of gene gRNA-group pairs: %d rows %d columns.\n', nrow(gene_gRNA_group_pairs), ncol(gene_gRNA_group_pairs)))

table(gene_gRNA_group_pairs$pair_type)
```

```{r run-sceptre-stimulated-data, eval=FALSE}
result <- run_sceptre_high_moi(gene_matrix = gene_matrix,
                               combined_perturbation_matrix = combined_perturbation_matrix,
                               covariate_matrix = covariate_matrix,
                               gene_gRNA_group_pairs = gene_gRNA_group_pairs,
                               side = "both",
                               storage_dir = outdir, 
                               full_output = FALSE)

outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/simulated_data/sceptre_output'
saveRDS(result, file.path(outdir, 'sceptre.result.rds'))
```

#### Unstimulated T cells
```{r load-unstimulated-input-data}
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/unsimulated_data'
load(file.path(datadir, "data.matrices.RData"))
gene_gRNA_group_pairs <- readRDS(file.path(datadir, "gene.gRNA.group.pairs.rds"))

cat(sprintf('Dimenstion of gene expression matrix: %d rows %d columns.\n', nrow(gene_matrix), ncol(gene_matrix)))
cat(sprintf('Dimenstion of combined perturbation matrix: %d rows %d columns.\n', nrow(combined_perturbation_matrix), ncol(combined_perturbation_matrix)))
cat(sprintf('Dimenstion of covariate matrix: %d rows %d columns.\n', nrow(covariate_matrix), ncol(covariate_matrix)))
cat(sprintf('Dimenstion of gene gRNA-group pairs: %d rows %d columns.\n', nrow(gene_gRNA_group_pairs), ncol(gene_gRNA_group_pairs)))

table(gene_gRNA_group_pairs$pair_type)
```

```{r run-sceptre-unstimulated-data, eval=FALSE}
result <- run_sceptre_high_moi(gene_matrix = gene_matrix,
                               combined_perturbation_matrix = combined_perturbation_matrix,
                               covariate_matrix = covariate_matrix,
                               gene_gRNA_group_pairs = gene_gRNA_group_pairs,
                               side = "both",
                               storage_dir = outdir, 
                               full_output = FALSE)

outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/unsimulated_data/sceptre_output'
saveRDS(result, file.path(outdir, 'sceptre.result.rds'))
```

### Analyze the results

#### Stimulated T cells

```{r load-stimulated-result}
outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/simulated_data/sceptre_output'
simulated_result <- readRDS(file.path(outdir, 'sceptre.result.rds'))
head(simulated_result, 10)
```

Negative control pairs
```{r neg-control-stimulated,  fig.width=5, fig.height=5}
neg_control_p_vals <- simulated_result %>% filter(pair_type == "negative_control") %>% pull(p_value)
qq_plot <- make_qq_plot(neg_control_p_vals)
plot(qq_plot)
```

Candidate pairs
```{r fdr-stimulated}
simulated_candidate_pair_results <- simulated_result %>% filter(pair_type == "candidate") %>%
  mutate(p_val_adj = p.adjust(p_value, method = "BH"))
```

```{r candidate-stimulated-qqplot, fig.width=5, fig.height=5}
simulated_candidate_pvals <- simulated_candidate_pair_results %>% pull(p_value)
make_qq_plot(simulated_candidate_pvals)
```

```{r select-sig-pairs-stimulated}
simulated_discovery_set <- simulated_candidate_pair_results %>% filter(p_val_adj <= 0.1)
head(simulated_discovery_set)
```

```{r}
saveRDS(simulated_candidate_pair_results, file.path(outdir, 'sceptre.candidate.pair.results.rds'))
saveRDS(simulated_discovery_set, file.path(outdir, 'sceptre.discovery.set.results.rds'))
```

#### Unstimulated T cells

```{r load-unstimulated-result}
outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_cropseq_data/unsimulated_data/sceptre_output'
unsimulated_result <- readRDS(file.path(outdir, 'sceptre.result.rds'))
head(unsimulated_result, 10)
```

Negative control pairs
```{r neg-control-unstimulated,  fig.width=5, fig.height=5}
neg_control_p_vals <- unsimulated_result %>% filter(pair_type == "negative_control") %>% pull(p_value)
qq_plot <- make_qq_plot(neg_control_p_vals)
plot(qq_plot)
```

Candidate pairs
```{r fdr-unstimulated}
unsimulated_candidate_pair_results <- unsimulated_result %>% filter(pair_type == "candidate") %>%
  mutate(p_val_adj = p.adjust(p_value, method = "BH"))
```

```{r candidate-unstimulated-qqplot, fig.width=5, fig.height=5}
unsimulated_candidate_pvals <- unsimulated_candidate_pair_results %>% pull(p_value)
make_qq_plot(unsimulated_candidate_pvals)
```

```{r select-sig-pairs-unstimulated}
unsimulated_discovery_set <- unsimulated_candidate_pair_results %>% filter(p_val_adj <= 0.1)
head(unsimulated_discovery_set)
```

```{r}
saveRDS(unsimulated_candidate_pair_results, file.path(outdir, 'sceptre.candidate.pair.results.rds'))
saveRDS(unsimulated_discovery_set, file.path(outdir, 'sceptre.discovery.set.results.rds'))
```


## Permutation analysis of stimulated data

Pool results from permutation data sets
```{r}
outdir <- "/project2/xinhe/kevinluo/GSFA/sceptre_analysis/TCells_data_updated/simulated_data/sceptre_output/permutation"
sceptre_res <- data.frame()
for(permute_num in 1:10){
  res_dir <- paste0(outdir, "/perm_", permute_num)
  res <- readRDS(file.path(res_dir, 'sceptre.result.rds'))
  sceptre_res <- rbind(sceptre_res, data.frame(permute_num = permute_num, res))
}
```

```{r permutation-qqplot, fig.width=5, fig.height=5}
candidate_p_vals <- sceptre_res %>% filter(pair_type == "candidate") %>% pull(p_value)
qqunif.plot(candidate_p_vals, main = "SCEPTRE (permutation)")
```

