---
title: "SCEPTRE analysis on LUHMES CROP-seq data"
author: "Kaixuan Luo"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

## About the data sets

CROP-seq datasets: `/project2/xinhe/yifan/Factor_analysis/shared_data/`.

The data are Seurat objects, with raw gene counts stored in `obj@assays$RNA@counts`, and cell meta data stored in `obj@meta.data`.
Normalized and scaled data used for GSFA are stored in `obj@assays$RNA@scale.data`, the rownames of which are the 6k genes used for GSFA.

## Analysis

Load packages
```{r, message=FALSE, warning=FALSE}
dyn.load('/software/geos-3.7.0-el7-x86_64/lib64/libgeos_c.so') # attach the geos lib for Seurat
suppressPackageStartupMessages(library(tidyverse))
library(cowplot)
library(Matrix)
library(sceptre)
library(Seurat)
source("R/plotting_functions.R")
```

### Prepare input data for SCEPTRE

Load the Seurat object of LUHMES data
```{r load-data}
LUHMES_data <- readRDS('/project2/xinhe/yifan/Factor_analysis/shared_data/LUHMES_cropseq_data_seurat.rds')
datadir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/LUHMES_cropseq_data'
```

We first prepare three separate data objects required to run SCEPTRE: 
the gene expression matrix, the perturbation matrix, and the covariate matrix.

1. Gene expression matrices
Gene expression (gene-by-cell) raw count matrix
```{r exp-matrix}
gene_matrix <- LUHMES_data@assays$RNA@counts
# gene-by-cell expression matrix
gene_matrix[1:10, 1:3]
dim(gene_matrix)
```

2. Cell meta data and covariate matrix
```{r cell-covariate-matrix}
metadata <- LUHMES_data@meta.data
metadata[1:5, ]
covariate_matrix <- metadata[,c('orig.ident', 'nCount_RNA', 'nFeature_RNA', 'percent_mt')]
covariate_matrix[1:5,]
dim(covariate_matrix)
```

3. Perturbation matrix (a binary matrix of perturbations,  rows are gRNA groups and columns are cell barcodes)
```{r perturbation-matrix}
combined_perturbation_matrix <- t(metadata[,4:18])
dim(combined_perturbation_matrix)
combined_perturbation_matrix[1:10,1:3]
range(combined_perturbation_matrix)
```

4. Specify the gene-gRNA group pairs to test for association

Include the 6k genes used for GSFA as candidates
```{r select-genes}
# Normalized and scaled data used for GSFA, the rownames of which are the 6k genes used for GSFA
scaled_gene_matrix <- LUHMES_data@assays$RNA@scale.data
dim(scaled_gene_matrix)
selected_gene_id <- rownames(scaled_gene_matrix)
all(selected_gene_id %in% rownames(gene_matrix))
```

```{r select-pairs}
gRNA_group <- rownames(combined_perturbation_matrix)
pairs <- expand.grid(selected_gene_id, gRNA_group)
gene_gRNA_group_pairs <- data.frame(gene_id = pairs$Var1, gRNA_group = pairs$Var2, pair_type = "candidate")
gene_gRNA_group_pairs[gene_gRNA_group_pairs$gRNA_group == "Nontargeting", "pair_type"] <- "negative_control"
table(gene_gRNA_group_pairs$pair_type)
table(gene_gRNA_group_pairs$gRNA_group)
dim(gene_gRNA_group_pairs)
```

```{r, eval=FALSE}
save(list = c("gene_matrix", "combined_perturbation_matrix", "covariate_matrix"),
     file = file.path(datadir, 'data.matrices.RData'))
saveRDS(gene_gRNA_group_pairs, file.path(datadir, "gene.gRNA.group.pairs.rds"))
```

### Analyze the SCEPTRE results

```{r load-result}
outdir <- '/project2/xinhe/kevinluo/GSFA/sceptre_analysis/LUHMES_data_updated/sceptre_output'
result <- readRDS(file.path(outdir, 'sceptre.result.rds'))
```

Negative control pairs
```{r negative-control-qqplot, fig.width=5, fig.height=5}
neg_control_p_vals <- result %>% filter(pair_type == "negative_control") %>% pull(p_value)
make_qq_plot(neg_control_p_vals)
```

Candidate pairs
```{r fdr}
candidate_pair_results <- result %>% filter(pair_type == "candidate") %>%
  mutate(p_val_adj = p.adjust(p_value, method = "BH"))
```

```{r candidate-qqplot, fig.width=5, fig.height=5}
candidate_pvals <- candidate_pair_results %>% pull(p_value)
make_qq_plot(candidate_pvals)
```

```{r select-sig-pairs}
discovery_set <- candidate_pair_results %>% filter(p_val_adj <= 0.1)
```

### Permutation analysis

Pool results from 10 permuted data sets
```{r}
outdir <- "/project2/xinhe/kevinluo/GSFA/sceptre_analysis/LUHMES_data_updated/sceptre_output/permutation"

sceptre_res <- data.frame()
for(permute_num in 1:10){
  res_dir <- paste0(outdir, "/perm_", permute_num)
  res <- readRDS(file.path(res_dir, 'sceptre.result.rds'))
  sceptre_res <- rbind(sceptre_res, data.frame(permute_num = permute_num, res))
}
```

```{r permutation-qqplot, fig.width=5, fig.height=5}
candidate_pair_results <- sceptre_res %>% filter(pair_type == "candidate") %>%
  mutate(p_val_adj = p.adjust(p_value, method = "BH"))

candidate_p_vals <- candidate_pair_results %>% pull(p_value)

qqunif.plot(candidate_p_vals, main = "SCEPTRE (permutation)")
```

