---
title: "Two-step clustering analysis on LUHMES CROP-seq data"
author: "Kaixuan Luo"
date: "2022-08-10"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(results = "hold", fig.align = "center", message=FALSE, warning=FALSE)
```

```{bash, include=FALSE, eval=FALSE}
mkdir -p /project2/xinhe/kevinluo/GSFA/data

cp /project2/xinhe/yifan/Factor_analysis/shared_data/LUHMES_cropseq_data_seurat.rds \
  /project2/xinhe/kevinluo/GSFA/data

cp /project2/xinhe/yifan/Factor_analysis/LUHMES/GSE142078_raw/GSM4219575_Run1_genes.tsv.gz \
  /project2/xinhe/kevinluo/GSFA/data/LUHMES_GSM4219575_Run1_genes.tsv.gz
```

Scripts for running the analysis: `code/run_twostep_clustering_LUHMES_data.R`

```{bash, include=FALSE, eval=FALSE}
sbatch --mem=40G ~/projects/GSFA_analysis/code/run_twostep_clustering_LUHMES_data.sbatch
```


## Load the data sets

CROP-seq datasets: `/project2/xinhe/yifan/Factor_analysis/shared_data/LUHMES_cropseq_data_seurat.rds`

The data are Seurat objects, with raw gene counts stored in `obj@assays$RNA@counts`, 
and cell meta data stored in `obj@meta.data`.
Normalized and scaled data used for GSFA are stored in `obj@assays$RNA@scale.data`, 
the rownames of which are the 6k genes used for GSFA.

Load packages
```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(data.table))
dyn.load('/software/geos-3.7.0-el7-x86_64/lib64/libgeos_c.so') # attach the geos lib for Seurat
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(ggplot2))
require(reshape2)
require(dplyr)
require(ComplexHeatmap)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 13),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
suppressPackageStartupMessages(library(gridExtra))
source("code/plotting_functions.R")
```

Set directories
```{r dir}
data_dir <- "/project2/xinhe/kevinluo/GSFA/data/"
res_dir <- "/project2/xinhe/kevinluo/GSFA/twostep_clustering/LUHMES/"
dir.create(res_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load two-step clustering analysis result

```{r}
combined_obj <- readRDS(file.path(res_dir, "LUHMES_seurat_clustered.rds"))
```

## Associate perturbations with clusters

```{r associate-perturbs-clusters}
perturb_matrix <- combined_obj@meta.data[, 4:18]

cluster_labels <- Idents(combined_obj)
cluster_labels <- as.factor(as.numeric(as.character(cluster_labels))+1)
new_cluster_labels <- paste0("k", levels(cluster_labels))
names(new_cluster_labels) <- levels(combined_obj)
combined_obj <- RenameIdents(combined_obj, new_cluster_labels)

cluster_matrix <- matrix(0, nrow = nrow(perturb_matrix), ncol = length(levels(cluster_labels)))
cluster_matrix[cbind(1:nrow(perturb_matrix), cluster_labels)] <- 1
rownames(cluster_matrix) <- rownames(perturb_matrix)
colnames(cluster_matrix) <- new_cluster_labels
```

Use Chi-squared tests for the association of perturbations and clusters (2 x 2 tables)
```{r chisq-test}
summary_df <- expand.grid(colnames(perturb_matrix), colnames(cluster_matrix))
colnames(summary_df) <- c("perturb", "cluster")

summary_df <- cbind(summary_df, statistic = NA, stdres = NA, pval = NA)

for(i in 1:nrow(summary_df)){
  dt <- table(data.frame(perturb = perturb_matrix[,summary_df$perturb[i]], 
                         cluster = cluster_matrix[,summary_df$cluster[i]]))
  chisq <- chisq.test(dt)
  summary_df[i, ]$statistic <- chisq$statistic
  summary_df[i, ]$stdres <- chisq$stdres[2,2]
  summary_df[i, ]$pval <- chisq$p.value
}

summary_df$fdr <- p.adjust(summary_df$pval, method = "BH")
summary_df$bonferroni_adj <- p.adjust(summary_df$pval, method = "bonferroni")

# saveRDS(summary_df, file = file.path(res_dir, "LUHMES_seurat_guide_cluster_chisq_summary_df.rds"))

stdres_mat <- reshape2::dcast(summary_df %>% dplyr::select(perturb, cluster, stdres), perturb ~ cluster, value.var = "stdres")
rownames(stdres_mat) <- stdres_mat$perturb
stdres_mat$perturb <- NULL

fdr_mat <- reshape2::dcast(summary_df %>% dplyr::select(perturb, cluster, fdr), perturb ~ cluster, value.var = "fdr")
rownames(fdr_mat) <- fdr_mat$perturb
fdr_mat$perturb <- NULL

bonferroni_mat <- reshape2::dcast(summary_df %>% dplyr::select(perturb, cluster, bonferroni_adj), 
                                  perturb ~ cluster, value.var = "bonferroni_adj")
rownames(bonferroni_mat) <- bonferroni_mat$perturb
bonferroni_mat$perturb <- NULL
```

```{r stdres-fdr-dotplot, eval=TRUE, fig.width=9, fig.height=5.5}
KO_names <- rownames(fdr_mat)
dotplot_effectsize(t(stdres_mat), t(fdr_mat),
                   reorder_markers = c(KO_names[KO_names!="Nontargeting"], "Nontargeting"),
                   color_lgd_title = "Chi-squared test\nstandardized residuals",
                   size_lgd_title = "FDR",
                   max_score = 4,
                   min_score = -4,
                   by_score = 2) + coord_flip()
```

Plot perturbation ~ cluster associations (show Bonferroni adjusted p-values)
```{r stdres-bonferroni-dotplot, eval=TRUE, fig.width=9, fig.height=5.5}
KO_names <- rownames(bonferroni_mat)
dotplot_effectsize(t(stdres_mat), t(bonferroni_mat),
                   reorder_markers = c(KO_names[KO_names!="Nontargeting"], "Nontargeting"),
                   color_lgd_title = "Chi-squared test\nstandardized residuals",
                   size_lgd_title = "Bonferroni\nadjusted p-value",
                   max_score = 4,
                   min_score = -4,
                   by_score = 2) + coord_flip()
```

## Find DE genes for each cluster and assign DE genes to associated perturbations

First, find DE genes for each cluster using MAST (Bonferroni adjusted p-values < 0.05),
Then, for each perturbation, find the associated clusters, and pull the DE genes for those clusters.

```{r get-de-genes}
feature.names <- data.frame(fread(file.path(data_dir, "LUHMES_GSM4219575_Run1_genes.tsv.gz"),
                                  header = FALSE), stringsAsFactors = FALSE)

de.markers <- readRDS(file.path(res_dir, "LUHMES_seurat_MAST_DEGs.rds"))
names(de.markers) <- paste0("k", levels(cluster_labels))

de.genes.clusters <- vector("list", length = length(de.markers))
names(de.genes.clusters) <- names(de.markers)
for( i in 1:length(de.genes.clusters)){
  de_sumstats <- de.markers[[i]]
  de_genes <- unique(rownames(de_sumstats[de_sumstats$p_val_adj < 0.05,]))
  # de_genes <- feature.names$V2[match(de_genes, feature.names$V1)]
  de.genes.clusters[[i]] <- de_genes
}
```


Number of DE genes for each perturbation (Chi-squared test FDR < 0.05)
```{r count-de-genes-fdr, fig.width=13, fig.height=5}
perturb_names <- colnames(perturb_matrix)
perturb_names <- c("Nontargeting", perturb_names[perturb_names!="Nontargeting"])

de.genes.perturbs <- vector("list", length = length(perturb_names))
names(de.genes.perturbs) <- perturb_names

for(i in 1:length(de.genes.perturbs)){
  perturb_name <- names(de.genes.perturbs)[i]
  associated_cluster_labels <- colnames(fdr_mat)[which(fdr_mat[perturb_name, ] < 0.05)]
  if(length(associated_cluster_labels) > 0){
    de.genes.perturbs[[i]] <- unique(unlist(de.genes.clusters[associated_cluster_labels]))
  }
}

num.de.genes.perturbs <- sapply(de.genes.perturbs, length)

twostep_clustering_fdr0.05_genes <- de.genes.perturbs

dge_plot_df <- data.frame(Perturbation = names(num.de.genes.perturbs), Num_DEGs = num.de.genes.perturbs)
dge_plot_df$Perturbation <- factor(dge_plot_df$Perturbation, levels = names(num.de.genes.perturbs))

ggplot(data=dge_plot_df, aes(x = Perturbation, y = Num_DEGs+1)) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_text(aes(label = Num_DEGs), position=position_dodge(width=0.9), vjust=-0.25) +
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Target gene",
       y = "Number of DEGs",
       title = "Number of DEGs detected by Two-step clustering with MAST DE analysis") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 13))
```

Number of DE genes for each perturbation (Chi-squared test Bonferroni adjusted p-value < 0.05)
```{r count-de-genes-bonferroni, fig.width=13, fig.height=5}

perturb_names <- colnames(perturb_matrix)
perturb_names <- c("Nontargeting", perturb_names[perturb_names!="Nontargeting"])

de.genes.perturbs <- vector("list", length = length(perturb_names))
names(de.genes.perturbs) <- perturb_names

for(i in 1:length(de.genes.perturbs)){
  perturb_name <- names(de.genes.perturbs)[i]
  associated_cluster_labels <- colnames(bonferroni_mat)[which(bonferroni_mat[perturb_name, ] < 0.05)]
  if(length(associated_cluster_labels) > 0){
    de.genes.perturbs[[i]] <- unique(unlist(de.genes.clusters[associated_cluster_labels]))
  }
}

num.de.genes.perturbs <- sapply(de.genes.perturbs, length)

twostep_clustering_bonferroni0.05_genes <- de.genes.perturbs

dge_plot_df <- data.frame(Perturbation = names(num.de.genes.perturbs), Num_DEGs = num.de.genes.perturbs)
dge_plot_df$Perturbation <- factor(dge_plot_df$Perturbation, levels = names(num.de.genes.perturbs))

ggplot(data=dge_plot_df, aes(x = Perturbation, y = Num_DEGs+1)) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_text(aes(label = Num_DEGs), position=position_dodge(width=0.9), vjust=-0.25) +
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Target gene",
       y = "Number of DEGs",
       title = "Number of DEGs detected by Two-step clustering with MAST DE analysis") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 13))
```


## Compare single-gene DE p-value distributions between two-step clustering analysis and GSFA

```{r}
fdr_cutoff <- 0.05
lfsr_cutoff <- 0.05
```

Load the output of GSFA `fit_gsfa_multivar()` run. 
```{r}
data_folder <- "/project2/xinhe/yifan/Factor_analysis/LUHMES/"
fit <- readRDS(paste0(data_folder,
                      "gsfa_output_detect_01/use_negctrl/All.gibbs_obj_k20.svd_negctrl.seed_14314.light.rds"))
gibbs_PM <- fit$posterior_means
lfsr_mat <- fit$lfsr[, -ncol(fit$lfsr)]
total_effect <- fit$total_effect[, -ncol(fit$total_effect)]
KO_names <- colnames(lfsr_mat)
```

DEGs detected by GSFA
```{r echo=FALSE}
gsfa_sig_genes <- apply(lfsr_mat, 2, function(x){names(x)[x < lfsr_cutoff]})

sapply(gsfa_sig_genes, length)
```

Load MAST single-gene DE result
```{r mast}
guides <- KO_names[KO_names!="Nontargeting"]

mast_list <- list()
for (m in guides){
  fname <- paste0(data_folder, "processed_data/MAST/dev_top6k_negctrl/gRNA_", m, ".dev_res_top6k.vs_negctrl.rds")
  tmp_df <- readRDS(fname)
  tmp_df$geneID <- rownames(tmp_df)
  tmp_df <- tmp_df %>% dplyr::rename(FDR = fdr, PValue = pval)
  mast_list[[m]] <- tmp_df
}
mast_signif_counts <- sapply(mast_list, function(x){filter(x, FDR < fdr_cutoff) %>% nrow()})
```

QQ-plots of MAST DE p-values for the GSFA significant genes vs two-step clustering DE genes.
```{r qqplots-gsfa-tsc-all, fig.width=12, fig.height=8}
qqplots <- list()
for(i in 1:length(guides)){
  guide <- guides[i]
  mast_res <- mast_list[[guide]]
  tsc_de_genes <- twostep_clustering_fdr0.05_genes[[guide]]
  gsfa_de_genes <- gsfa_sig_genes[[guide]]
  tsc_de_genes <- intersect(tsc_de_genes, rownames(mast_res))
  gsfa_de_genes <- intersect(gsfa_de_genes, rownames(mast_res))
  
  if(length(tsc_de_genes)>0 && length(gsfa_de_genes) >0){
    mast_res$tsc_gene <- 0
    mast_res[tsc_de_genes, ]$tsc_gene <- 1
    mast_res$gsfa_gene <- 0
    mast_res[gsfa_de_genes, ]$gsfa_gene <- 1
    pvalue_list <- list('Two-step clustering'=dplyr::filter(mast_res,tsc_gene==1)$PValue,
                        'GSFA'=dplyr::filter(mast_res,gsfa_gene==1)$PValue,
                        'all genes'=mast_res$PValue)
    qqplots[[guide]] <- qqplot.pvalue(pvalue_list, pointSize = 1, legendSize = 4) + 
      ggtitle(guide) + theme(plot.title = element_text(hjust = 0.5)) +
      scale_colour_discrete(name="Method")
  }
}

grid.arrange(grobs = qqplots, nrow = 2, ncol = 2)
```

Pooling p-values from all perturbations
```{r qqplots-gsfa-tsc-all-combined, dev=c('png', 'pdf'), fig.width=8, fig.height=6}
combined_mast_res <- data.frame()
for(i in 1:length(guides)){
  guide <- guides[i]
  mast_res <- mast_list[[guide]]
  tsc_de_genes <- twostep_clustering_fdr0.05_genes[[guide]]
  gsfa_de_genes <- gsfa_sig_genes[[guide]]
  tsc_de_genes <- intersect(tsc_de_genes, rownames(mast_res))
  gsfa_de_genes <- intersect(gsfa_de_genes, rownames(mast_res))
  mast_res$tsc_gene <- 0
  if(length(tsc_de_genes) >0){
    mast_res[tsc_de_genes, ]$tsc_gene <- 1
  }
  mast_res$gsfa_gene <- 0
  if(length(gsfa_de_genes) >0){
    mast_res[gsfa_de_genes, ]$gsfa_gene <- 1
  }
  combined_mast_res <- rbind(combined_mast_res, mast_res)
}

pvalue_list <- list('Two-step clustering'=dplyr::filter(combined_mast_res,tsc_gene==1)$PValue,
                    'GSFA'=dplyr::filter(combined_mast_res,gsfa_gene==1)$PValue,
                    'all genes'=combined_mast_res$PValue)

qqplot.pvalue(pvalue_list, pointSize = 1, legendSize = 4) + 
      ggtitle("") + theme(plot.title = element_text(hjust = 0.5)) +
      scale_colour_discrete(name="Method")
```

QQ-plots of MAST DE p-values for the GSFA only genes vs two-step only genes.
```{r qqplots-gsfa-tsc-only, fig.width=12, fig.height=8}
qqplots <- list()
for(i in 1:length(guides)){
  guide <- guides[i]
  mast_res <- mast_list[[guide]]
  tsc_de_genes <- twostep_clustering_fdr0.05_genes[[guide]]
  gsfa_de_genes <- gsfa_sig_genes[[guide]]
  tsc_de_genes <- intersect(tsc_de_genes, rownames(mast_res))
  gsfa_de_genes <- intersect(gsfa_de_genes, rownames(mast_res))
  
  if(length(tsc_de_genes)>0 && length(gsfa_de_genes) >0){
    mast_res$tsc_only_gene <- 0
    mast_res[setdiff(tsc_de_genes, gsfa_de_genes), ]$tsc_only_gene <- 1
    mast_res$gsfa_only_gene <- 0
    mast_res[setdiff(gsfa_de_genes, tsc_de_genes), ]$gsfa_only_gene <- 1
    
    pvalue_list <- list('Two-step clustering only'=dplyr::filter(mast_res,tsc_only_gene==1)$PValue,
                        'GSFA only'=dplyr::filter(mast_res,gsfa_only_gene==1)$PValue,
                        'all genes'=mast_res$PValue)
    qqplots[[guide]] <- qqplot.pvalue(pvalue_list, pointSize = 1, legendSize = 4) + 
      ggtitle(guide) + theme(plot.title = element_text(hjust = 0.5)) +
      scale_colour_discrete(name="Method")

  }
}

grid.arrange(grobs = qqplots, nrow = 2, ncol = 2)
```

Pooling p-values from all perturbations
```{r qqplots-gsfa-tsc-only-combined, dev=c('png', 'pdf'), fig.width=8, fig.height=6}
combined_mast_res <- data.frame()
for(i in 1:length(guides)){
  guide <- guides[i]
  mast_res <- mast_list[[guide]]
  tsc_de_genes <- twostep_clustering_fdr0.05_genes[[guide]]
  gsfa_de_genes <- gsfa_sig_genes[[guide]]
  tsc_de_genes <- intersect(tsc_de_genes, rownames(mast_res))
  gsfa_de_genes <- intersect(gsfa_de_genes, rownames(mast_res))
  mast_res$tsc_only_gene <- 0
  if(length(setdiff(tsc_de_genes, gsfa_de_genes)) >0){
    mast_res[setdiff(tsc_de_genes, gsfa_de_genes), ]$tsc_only_gene <- 1
  }
  mast_res$gsfa_only_gene <- 0
  if(length(setdiff(gsfa_de_genes, tsc_de_genes)) >0){
    mast_res[setdiff(gsfa_de_genes, tsc_de_genes), ]$gsfa_only_gene <- 1
  }
  combined_mast_res <- rbind(combined_mast_res, mast_res)
}

pvalue_list <- list('Two-step clustering only'=dplyr::filter(combined_mast_res,tsc_only_gene==1)$PValue,
                    'GSFA only'=dplyr::filter(combined_mast_res,gsfa_only_gene==1)$PValue,
                    'all genes'=combined_mast_res$PValue)

qqplot.pvalue(pvalue_list, pointSize = 1, legendSize = 4) + 
      ggtitle("") + theme(plot.title = element_text(hjust = 0.5)) +
      scale_colour_discrete(name="Method")
```

