---
title: "Unguided GSFA analysis on LUHMES CROP-seq data"
author: "Kaixuan Luo"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(results = "hold", fig.align = "center", message=FALSE, warning=FALSE)
```

Load packages
```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(data.table))
# suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(ggplot2))
require(reshape2)
require(dplyr)
require(ComplexHeatmap)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 13),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
suppressPackageStartupMessages(library(gridExtra))
source("R/plotting_functions.R")
```

Set directories
```{r dir}
res_dir <- "/project2/xinhe/kevinluo/GSFA/unguided_GSFA/LUHMES/"
dir.create(res_dir, recursive = TRUE, showWarnings = FALSE)
```

Load unguided GSFA result
```{r}
fit <- readRDS("/project2/xinhe/kevinluo/GSFA/unguided_GSFA/LUHMES/unguided_gsfa_output/All.gibbs_obj_k20.unguided.svd.seed_14314.light.rds")
```

Load the cell by perturbation matrix.
```{r}
data_folder <- "/project2/xinhe/kevinluo/GSFA/unguided_GSFA/LUHMES/"
metadata <- readRDS(paste0(data_folder, "processed_data/merged_metadata.rds"))
# Perturbation info:
G_mat <- metadata[, 4:18]
G_mat <- as.matrix(G_mat)
KO_names <- colnames(G_mat)
negctrl_index <- which(KO_names == "Nontargeting")
```

Use linear regression to test for the association between perturbations and factors
```{r chisq-test}
Z_pm <- fit$posterior_means$Z_pm
if(!all.equal(rownames(G_mat), rownames(Z_pm))){
  stop("Rownames of G_mat do not match with Z_pm!")
}

perturb_matrix <- G_mat
factor_matrix <- Z_pm

summary_df <- expand.grid(colnames(perturb_matrix), colnames(factor_matrix))
colnames(summary_df) <- c("perturb", "factor")

summary_df <- cbind(summary_df, beta = NA, statistic = NA, pval = NA)

for(i in 1:nrow(summary_df)){
  df <- data.frame(perturb = perturb_matrix[,summary_df$perturb[i]], 
                   factor = factor_matrix[,summary_df$factor[i]])
  lm.res <- lm(factor ~ perturb, data=df)
  summary_df[i, ]$beta <- summary(lm.res)$coefficients["perturb",1]
  summary_df[i, ]$statistic <- summary(lm.res)$coefficients["perturb",3]
  summary_df[i, ]$pval <- summary(lm.res)$coefficients["perturb",4]
}

summary_df$fdr <- p.adjust(summary_df$pval, method = "BH")
summary_df$bonferroni_adj <- p.adjust(summary_df$pval, method = "bonferroni")

saveRDS(summary_df, file = file.path(res_dir, "LUHMES_unguidedGSFA_guide_factor_lm_summary_df.rds"))

stat_mat <- reshape2::dcast(summary_df %>% dplyr::select(perturb, factor, statistic), perturb ~ factor, value.var = "statistic") 
rownames(stat_mat) <- stat_mat$perturb
stat_mat$perturb <- NULL
stat_mat <- as.matrix(stat_mat)

beta_mat <- reshape2::dcast(summary_df %>% dplyr::select(perturb, factor, beta), perturb ~ factor, value.var = "beta")
rownames(beta_mat) <- beta_mat$perturb
beta_mat$perturb <- NULL
beta_mat <- as.matrix(beta_mat)

fdr_mat <- reshape2::dcast(summary_df %>% dplyr::select(perturb, factor, fdr), perturb ~ factor, value.var = "fdr")
rownames(fdr_mat) <- fdr_mat$perturb
fdr_mat$perturb <- NULL
fdr_mat <- as.matrix(fdr_mat)

bonferroni_mat <- reshape2::dcast(summary_df %>% dplyr::select(perturb, factor, bonferroni_adj), 
                                  perturb ~ factor, value.var = "bonferroni_adj")
rownames(bonferroni_mat) <- bonferroni_mat$perturb
bonferroni_mat$perturb <- NULL
bonferroni_mat <- as.matrix(bonferroni_mat)

```

```{r stat-fdr-dotplot, eval=TRUE, fig.width=9, fig.height=5.5}
# pdf(file.path(res_dir, "stat-fdr-dotplot.pdf"), width = 9, height = 5.5)
KO_names <- rownames(fdr_mat)
dotplot_effectsize(t(beta_mat), t(fdr_mat),
                   reorder_markers = c(KO_names[KO_names!="Nontargeting"], "Nontargeting"),
                   color_lgd_title = "Estimated effect size",
                   size_lgd_title = "FDR",
                   max_score = 0.6,
                   min_score = -0.6,
                   by_score = 0.3) + coord_flip()

# dev.off()
```

Plot perturbation ~ cluster associations (show Bonferroni adjusted p-values)
```{r stat-bonferroni-dotplot, eval=TRUE, fig.width=9, fig.height=5.5}
# pdf(file.path(res_dir, "stat-bonferroni-dotplot.pdf"), width = 9, height = 5.5)
KO_names <- rownames(bonferroni_mat)
dotplot_effectsize(t(beta_mat), t(bonferroni_mat),
                   reorder_markers = c(KO_names[KO_names!="Nontargeting"], "Nontargeting"),
                   color_lgd_title = "Estimated effect size",
                   size_lgd_title = "Bonferroni\nadjusted p-value",
                   max_score = 0.6,
                   min_score = -0.6,
                   by_score = 0.3) + coord_flip()
# dev.off()
```

## Find DE genes for each factor and assign DE genes to associated perturbations

First, find DE genes for each factor using F matrix (PIP>0.95).

Then, for each perturbation, find the associated factors, and pull the DE genes for those factors.

```{r get-de-genes}
F_pm <- fit$posterior_means$F_pm
# dim(F_pm)
# feature.names <- data.frame(fread(file.path(data_dir, "LUHMES_GSM4219575_Run1_genes.tsv.gz"),
#                                   header = FALSE), stringsAsFactors = FALSE)

de.genes.factors <- vector("list", length = ncol(F_pm))
names(de.genes.factors) <- colnames(F_pm)
for( i in 1:length(de.genes.factors)){
  de_genes <- rownames(F_pm[F_pm[,i] > 0.95,])
  # de_genes <- feature.names$V2[match(de_genes, feature.names$V1)]
  de.genes.factors[[i]] <- de_genes
}
```


Number of DE genes for each perturbation (FDR < 0.05)
```{r count-de-genes-fdr, fig.width=13, fig.height=5}
perturb_names <- colnames(perturb_matrix)
perturb_names <- c("Nontargeting", perturb_names[perturb_names!="Nontargeting"])

de.genes.perturbs <- vector("list", length = length(perturb_names))
names(de.genes.perturbs) <- perturb_names

for(i in 1:length(de.genes.perturbs)){
  perturb_name <- names(de.genes.perturbs)[i]
  associated_factors <- colnames(fdr_mat)[which(fdr_mat[perturb_name, ] < 0.05)]
  if(length(associated_factors) > 0){
    de.genes.perturbs[[i]] <- unique(unlist(de.genes.factors[associated_factors]))
  }
}

num.de.genes.perturbs <- sapply(de.genes.perturbs, length)

unguided_GSFA_fdr0.05_genes <- de.genes.perturbs

dge_plot_df <- data.frame(Perturbation = names(num.de.genes.perturbs), Num_DEGs = num.de.genes.perturbs)
dge_plot_df$Perturbation <- factor(dge_plot_df$Perturbation, levels = names(num.de.genes.perturbs))

# pdf(file.path(res_dir, "count-de-genes.pdf"), width = 13, height = 5)
ggplot(data=dge_plot_df, aes(x = Perturbation, y = Num_DEGs+1)) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_text(aes(label = Num_DEGs), position=position_dodge(width=0.9), vjust=-0.25) +
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Target gene",
       y = "Number of DEGs",
       title = "Number of DEGs detected by unguided GSFA") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 13))
# dev.off()
```


Number of DE genes for each perturbation (Bonferroni adjusted p-value < 0.05)
```{r count-de-genes-bonferroni, fig.width=13, fig.height=5}

perturb_names <- colnames(perturb_matrix)
perturb_names <- c("Nontargeting", perturb_names[perturb_names!="Nontargeting"])

de.genes.perturbs <- vector("list", length = length(perturb_names))
names(de.genes.perturbs) <- perturb_names

for(i in 1:length(de.genes.perturbs)){
  perturb_name <- names(de.genes.perturbs)[i]
  associated_factors <- colnames(bonferroni_mat)[which(bonferroni_mat[perturb_name, ] < 0.05)]
  if(length(associated_factors) > 0){
    de.genes.perturbs[[i]] <- unique(unlist(de.genes.factors[associated_factors]))
  }
}

num.de.genes.perturbs <- sapply(de.genes.perturbs, length)
unguided_GSFA_bonferroni0.05_genes <- de.genes.perturbs

dge_plot_df <- data.frame(Perturbation = names(num.de.genes.perturbs), Num_DEGs = num.de.genes.perturbs)
dge_plot_df$Perturbation <- factor(dge_plot_df$Perturbation, levels = names(num.de.genes.perturbs))

ggplot(data=dge_plot_df, aes(x = Perturbation, y = Num_DEGs+1)) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_text(aes(label = Num_DEGs), position=position_dodge(width=0.9), vjust=-0.25) +
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Target gene",
       y = "Number of DEGs",
       title = "Number of DEGs detected by unguided GSFA") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 13))
```

## Compare single-gene DE p-value distributions between GSFA and unguided GSFA

```{r}
fdr_cutoff <- 0.05
lfsr_cutoff <- 0.05
```

Load the output of GSFA `fit_gsfa_multivar()` run. 
```{r}
data_folder <- "/project2/xinhe/yifan/Factor_analysis/LUHMES/"
fit <- readRDS(paste0(data_folder,
                      "gsfa_output_detect_01/use_negctrl/All.gibbs_obj_k20.svd_negctrl.seed_14314.light.rds"))
gibbs_PM <- fit$posterior_means
lfsr_mat <- fit$lfsr[, -ncol(fit$lfsr)]
total_effect <- fit$total_effect[, -ncol(fit$total_effect)]
KO_names <- colnames(lfsr_mat)
```

DEGs detected by GSFA
```{r echo=FALSE}
gsfa_sig_genes <- apply(lfsr_mat, 2, function(x){names(x)[x < lfsr_cutoff]})

sapply(gsfa_sig_genes, length)
```

Load MAST single-gene DE result
```{r mast}
guides <- KO_names[KO_names!="Nontargeting"]

mast_list <- list()
for (m in guides){
  fname <- paste0(data_folder, "processed_data/MAST/dev_top6k_negctrl/gRNA_", m, ".dev_res_top6k.vs_negctrl.rds")
  tmp_df <- readRDS(fname)
  tmp_df$geneID <- rownames(tmp_df)
  tmp_df <- tmp_df %>% dplyr::rename(FDR = fdr, PValue = pval)
  mast_list[[m]] <- tmp_df
}
mast_signif_counts <- sapply(mast_list, function(x){filter(x, FDR < fdr_cutoff) %>% nrow()})
# summary(mast_list)
```

QQ-plots of MAST DE p-values for the GSFA genes vs unguided GSFA genes.
```{r qqplots-gsfa-unguided-all, fig.width=12, fig.height=12, message=FALSE}
qqplots <- list()
for(i in 1:length(guides)){
  guide <- guides[i]
  mast_res <- mast_list[[guide]]
  unguided_gsfa_de_genes <- unguided_GSFA_fdr0.05_genes[[guide]]
  gsfa_de_genes <- gsfa_sig_genes[[guide]]
  unguided_gsfa_de_genes <- intersect(unguided_gsfa_de_genes, rownames(mast_res))
  gsfa_de_genes <- intersect(gsfa_de_genes, rownames(mast_res))
  
  if(length(unguided_gsfa_de_genes)>0 && length(gsfa_de_genes) >0){
    mast_res$unguided_gsfa_gene <- 0
    mast_res[unguided_gsfa_de_genes, ]$unguided_gsfa_gene <- 1
    mast_res$gsfa_gene <- 0
    mast_res[gsfa_de_genes, ]$gsfa_gene <- 1
    pvalue_list <- list('Unguided GSFA'=dplyr::filter(mast_res,unguided_gsfa_gene==1)$PValue,
                        'GSFA'=dplyr::filter(mast_res,gsfa_gene==1)$PValue,
                        'all genes'=mast_res$PValue)
    qqplots[[guide]] <- qqplot.pvalue(pvalue_list, pointSize = 1, legendSize = 4) + 
      ggtitle(guide) + theme(plot.title = element_text(hjust = 0.5))
  }
}

grid.arrange(grobs = qqplots, nrow = 4, ncol = 2)
```


Pooling p-values from all perturbations
```{r qqplots-gsfa-unguided-all-combined, dev=c('png', 'pdf'), fig.width=7.5, fig.height=6}
combined_mast_res <- data.frame()
for(i in 1:length(guides)){
  guide <- guides[i]
  mast_res <- mast_list[[guide]]
  unguided_gsfa_de_genes <- unguided_GSFA_fdr0.05_genes[[guide]]
  gsfa_de_genes <- gsfa_sig_genes[[guide]]
  unguided_gsfa_de_genes <- intersect(unguided_gsfa_de_genes, rownames(mast_res))
  gsfa_de_genes <- intersect(gsfa_de_genes, rownames(mast_res))
  mast_res$unguided_gsfa_gene <- 0
  if(length(unguided_gsfa_de_genes) >0){
    mast_res[unguided_gsfa_de_genes, ]$unguided_gsfa_gene <- 1
  }
  mast_res$gsfa_gene <- 0
  if(length(gsfa_de_genes) >0){
    mast_res[gsfa_de_genes, ]$gsfa_gene <- 1
  }
  combined_mast_res <- rbind(combined_mast_res, mast_res)
}

pvalue_list <- list('Unguided GSFA'=dplyr::filter(combined_mast_res,unguided_gsfa_gene==1)$PValue,
                    'GSFA'=dplyr::filter(combined_mast_res,gsfa_gene==1)$PValue,
                    'all genes'=combined_mast_res$PValue)

# pdf(file.path(res_dir, "qqplot_all_combined.pdf"))
qqplot.pvalue(pvalue_list, pointSize = 1, legendSize = 4) + 
      ggtitle("") + theme(plot.title = element_text(hjust = 0.5)) +
      scale_colour_discrete(name="Method")

# dev.off()
```

QQ-plots of MAST DE p-values for the GSFA only genes vs two-step only genes.
```{r qqplots-gsfa-unguided-only, fig.height=12, fig.width=12, message=FALSE}
qqplots <- list()
for(i in 1:length(guides)){
  guide <- guides[i]
  mast_res <- mast_list[[guide]]
  unguided_gsfa_de_genes <- unguided_GSFA_fdr0.05_genes[[guide]]
  gsfa_de_genes <- gsfa_sig_genes[[guide]]
  unguided_gsfa_de_genes <- intersect(unguided_gsfa_de_genes, rownames(mast_res))
  gsfa_de_genes <- intersect(gsfa_de_genes, rownames(mast_res))
  
  if(length(unguided_gsfa_de_genes)>0 && length(gsfa_de_genes) >0){
    mast_res$unguided_gsfa_only_gene <- 0
    mast_res[setdiff(unguided_gsfa_de_genes, gsfa_de_genes), ]$unguided_gsfa_only_gene <- 1
    mast_res$gsfa_only_gene <- 0
    mast_res[setdiff(gsfa_de_genes, unguided_gsfa_de_genes), ]$gsfa_only_gene <- 1
    
    pvalue_list <- list('Unguided GSFA only'=dplyr::filter(mast_res,unguided_gsfa_only_gene==1)$PValue,
                        'GSFA only'=dplyr::filter(mast_res,gsfa_only_gene==1)$PValue,
                        'all genes'=mast_res$PValue)
    qqplots[[guide]] <- qqplot.pvalue(pvalue_list, pointSize = 1, legendSize = 4) + 
      ggtitle(guide) + theme(plot.title = element_text(hjust = 0.5))
  }
}

grid.arrange(grobs = qqplots, nrow = 4, ncol = 2)
```


Pooling p-values from all perturbations
```{r qqplots-gsfa-unguided-only-combined, dev=c('png', 'pdf'), fig.width=7.5, fig.height=6}
combined_mast_res <- data.frame()
for(i in 1:length(guides)){
  guide <- guides[i]
  mast_res <- mast_list[[guide]]
  unguided_gsfa_de_genes <- unguided_GSFA_fdr0.05_genes[[guide]]
  gsfa_de_genes <- gsfa_sig_genes[[guide]]
  unguided_gsfa_de_genes <- intersect(unguided_gsfa_de_genes, rownames(mast_res))
  gsfa_de_genes <- intersect(gsfa_de_genes, rownames(mast_res))
  mast_res$unguided_gsfa_only_gene <- 0
  if(length(setdiff(unguided_gsfa_de_genes, gsfa_de_genes)) >0){
    mast_res[setdiff(unguided_gsfa_de_genes, gsfa_de_genes), ]$unguided_gsfa_only_gene <- 1
  }
  mast_res$gsfa_only_gene <- 0
  if(length(setdiff(gsfa_de_genes, unguided_gsfa_de_genes)) >0){
    mast_res[setdiff(gsfa_de_genes, unguided_gsfa_de_genes), ]$gsfa_only_gene <- 1
  }
  combined_mast_res <- rbind(combined_mast_res, mast_res)
}

pvalue_list <- list('Unguided GSFA only'=dplyr::filter(combined_mast_res,unguided_gsfa_only_gene==1)$PValue,
                    'GSFA only'=dplyr::filter(combined_mast_res,gsfa_only_gene==1)$PValue,
                    'all genes'=combined_mast_res$PValue)

# pdf(file.path(res_dir, "qqplot_only_combined.pdf"))
qqplot.pvalue(pvalue_list, pointSize = 1, legendSize = 4) + 
      ggtitle("") + theme(plot.title = element_text(hjust = 0.5)) +
      scale_colour_discrete(name="Method")
# dev.off()
```
